---
title: 'Lab 2'
author: "Kayla W., Dera C., Pavan E., Greg T."
output: pdf_document
toc: true
---
```{r load packages, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2) 
library(tidyverse)
library(haven)
library(patchwork)
library(kableExtra)
theme_set(theme_minimal())
install.packages('rstatix')
library(rstatix)
library(wooldridge)
library(car)
library(lmtest)
library(sandwich)
library(stargazer)
install.packages("data.table")
library("data.table")
library(zoo)
install.packages('psych')
library(psych)
library(gridExtra)
install.packages("lubridate")
library(lubridate)
```
# Pull data from [OWID Vaccination Github](https://github.com/owid/covid-19-data/tree/master/public/data/vaccinations)
```{r}
us_state_vaccinations <- fread("https://github.com/owid/covid-19-data/raw/master/public/data/vaccinations/us_state_vaccinations.csv")
head(arrange(us_state_vaccinations,desc(date)), n = 70)
```
# Pull in State Abbreviations
```{r}
states_abb <- tibble(state = state.name) %>%
   bind_cols(tibble(abb = state.abb)) %>% 
   bind_rows(tibble(state = "District of Columbia", abb = "DC"))
states_abb

?tibble()
```
## Prepare vaccination data
```{r}
# Convert date strings to dates. 
us_state_vaccinations[,date:=as.Date(date)]

# Select relevant Vacc. variables. 
vacc <- us_state_vaccinations %>%
  dplyr::select(date, location, people_vaccinated_per_hundred, people_fully_vaccinated_per_hundred) %>% rename(state = location) 

# Add state abbreviation
vacc <- left_join(vacc, states_abb, by = "state")

# Remove rows from locations that are not from 50 states + DC
vacc <- vacc[!is.na(vacc$abb), ]

# Create an effective date (14 days after vacc. date) to use for comparison with COVID cases
vacc <- vacc %>% 
  mutate(effective_date = date + 14) %>%
  rename(vacc_date = date) %>% 
  relocate(effective_date, vacc_date) %>% 
  filter(effective_date <= today()-1)

head(arrange(vacc,effective_date), n = 70)
```
# Vaccination data summary stats
```{r}
# review prepared dataset
describe(vacc)
summary(vacc)
```
## Pull data from cdc API [United States COVID-19 Cases and Deaths by State over Time](https://data.cdc.gov/Case-Surveillance/United-States-COVID-19-Cases-and-Deaths-by-State-o/9mfq-cb36) 
```{r}
# Load the package required to read JSON files.
library("rjson")

#install.packages("jsonlite")
library("jsonlite")
US_COVID_19_Cases_and_Deaths_by_State_over_Time <- jsonlite::fromJSON("https://data.cdc.gov/resource/9mfq-cb36.json?$limit=50000")

# Convert JSON file to a data frame.
cdc_covid_raw <- as.data.frame(US_COVID_19_Cases_and_Deaths_by_State_over_Time)
cdc_covid_raw %>% 
  arrange(state) %>% 
  group_by(state) %>% 
  arrange(desc(submission_date)) %>% 
  head(n=70)
```
# Prepare CDC data
```{r}
cdc_covid <- cdc_covid_raw

# Convert date strings to dates
cdc_covid$submission_date <- as.Date(cdc_covid$submission_date)

# Select relevant covid variables. 
cdc_covid <- cdc_covid %>%
  dplyr::select(submission_date, state, new_case, new_death)

# Change date column name
cdc_covid <- rename(cdc_covid, effective_date = submission_date, abb = state)

# Add state abbreviation
cdc_covid <- left_join(cdc_covid, states_abb, by = "abb")

# Remove rows from locations that are not from 50 states + DC
cdc_covid <- cdc_covid[!is.na(cdc_covid$state), ]

# Filter data to be only after first Vaccination effective date 01/26/2021
cdc_covid <- cdc_covid %>% filter(effective_date >= "2021-01-26")

# convert new_case and new_death to integers
cdc_covid$new_case <- as.numeric(as.character(cdc_covid$new_case))
cdc_covid$new_death <- as.numeric(as.character(cdc_covid$new_death))
```
# cdc_covid data summary stats
```{r}
# review prepared dataset
describe(cdc_covid)
summary(cdc_covid)
```
# Check how many rows have negative new_case or negative new_death
```{r}
head(arrange(cdc_covid, new_case), n = 20)
head(arrange(cdc_covid, new_death), n = 40)
```
# Mutate the negative new cases and negative new deaths to 0
```{r}
cdc_covid <- cdc_covid %>% mutate(new_case = replace(new_case, which(new_case<0), 0))
cdc_covid <- cdc_covid %>% mutate(new_death = replace(new_death, which(new_death<0), 0))

# Check that there are no more new cases or new deaths below 0
head(arrange(cdc_covid, new_case), n = 20)
head(arrange(cdc_covid, new_death), n = 40)
```
## Merge Vacc data with CDC COVID data
```{r}
vacc_cdccovid <- vacc %>% right_join(cdc_covid, by=c("effective_date","state"))

vacc_cdccovid <- vacc_cdccovid[, c("state", "vacc_date", "effective_date", "people_vaccinated_per_hundred", "people_fully_vaccinated_per_hundred", "new_case", "new_death")]

vacc_cdccovid %>%  arrange(desc(effective_date))
```
# Check summary stats on merged dataset
```{r}
describe(vacc_cdccovid)
summary(vacc_cdccovid)
```
# Time series of % pop vaccinated over time
```{r}
plot_vacc_1 <- vacc_cdccovid %>%  
  ggplot(aes(x = vacc_date, y = people_vaccinated_per_hundred)) + 
  geom_line(aes(color = state))  +
  labs(title = "Vaccionations are rising!", x = "Vaccination dates", y = "People Vaccinated per hundred") + 
  theme(plot.title = element_text(size=10))

plot_vacc_1
```
# Time series of % pop vaccinated over time by Effective Date
```{r}
plot_vacc_2 <- vacc_cdccovid %>%  
  ggplot(aes(x = effective_date, y = people_vaccinated_per_hundred)) + 
  geom_line(aes(color = state))  +
  labs(title = "Vaccionations are rising!", x = "Effective dates", y = "People Vaccinated per hundred") + 
  theme(plot.title = element_text(size=10))

plot_vacc_2
```
```{r}
vacc_cdccovid %>% filter(effective_date == "2021-01-02")
```


```{r}
plot_vacc_3 <- vacc_cdccovid %>%  
  ggplot(aes(x=people_vaccinated_per_hundred)) +
  geom_histogram(data=subset(vacc_cdccovid,effective_date == "2021-02-05"),fill = "red", alpha = 0.2) +
  geom_histogram(data=subset(vacc_cdccovid,effective_date == "2021-03-05"),fill = "blue", alpha = 0.2) +
  geom_histogram(data=subset(vacc_cdccovid,effective_date == "2021-04-05"),fill = "green", alpha = 0.2) +
  labs(title = "Histograms of People Vacc per hundred on effective dates: 2/05 (red), 3/05 (blue), 4/05 (green)", x = "People Vaccinated per hundred", y = "Counts") + 
  theme(plot.title = element_text(size=10))

plot_vacc_3
```
# Create cumulative sum of new cases and new deaths
```{r}
#install.packages("slider")
library(slider)
vacc_cdccovid <- vacc_cdccovid %>% 
  group_by(state) %>% 
  mutate(cumuluative_new_case_7 = slide_dbl(new_case, sum, .before = 6, .complete = TRUE)) %>% mutate(cumuluative_new_death_7 = slide_dbl(new_death, sum, .before = 6, .complete = TRUE))

# rearrange columns to have cumulative sum after daily count for new cases and new deaths
vacc_cdccovid <- vacc_cdccovid[, c("state", "vacc_date", "effective_date", "people_vaccinated_per_hundred", "people_fully_vaccinated_per_hundred", "new_case", "cumuluative_new_case_7", "new_death", "cumuluative_new_death_7")]

vacc_cdccovid
```
# add column of state population and then mutat new cases and new deaths to new_cases_per_100,000 and new_deaths_per_100,000

# [Census burea data](https://walker-data.com/tidycensus/articles/other-datasets.html)

# Change Plot below to have y axis be cumulative_new_case_7_per_100,000
```{r}
plot_1 <- vacc_cdccovid %>%  
  ggplot(aes(x = people_vaccinated_per_hundred, y = cumuluative_new_case_7)) + 
  geom_point()
plot_1
```


